Design and implementaiton of a live coding environment for data science

### 2017-11-16 live/paper.tex
Type checking text

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.3,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\section{Formalising live coding environment}
\label{sec:formal}

In this section, we present a formalisation of a live coding environment for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section.

@@

Finally, the evaluation of function values is similar, but requires three cases. If the body can 
be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
returns the value. If the body requires only the bound variable (\emph{fun-bind}), we return a 
lambda function with the delayed preview as the body. If the body requires further variables,
the result is a delayed preview containing a function.

\paragraph{Caching previews.}
For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews 
are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
$\bndclr{v} \rightarrow p$ from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
Whenever the $\Downarrow$ relation is used to obtain a preview for a graph node, we first 
attempt to find an already evaluated preview using the lookup table. If the preview has not
been previously evaluated, we evaluate it and add it to the lookup table.

@@

\begin{figure*}[t]
\begin{equation*}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma, \Delta, \bot}(e_0)\\
\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\end{array}
\qquad\qquad
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma_1, \Delta, \bot}(e)\\
\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
\\[0.75em]
\end{array}
\end{equation*}
\caption{Revised binding rules, tracking call sites of function values.} 
\label{fig:binding-rules-callsite}
\end{figure*}

\begin{figure*}
\begin{equation*}
\begin{array}{l}
\inference[(val)~]
  {\Sigma(n) = \tau}
  {\bnd{val}(n) \vdash \tau }
\\
\\
\inference[(var)~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E \\
    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\
\\
\inference[(fun-val)~]
  { (\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}) \in E \\
    (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i)) \in E \\
    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) \\
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\\
\\
\inference[(mem)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E \\
    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\  
\\
\end{array}
\end{equation*}

\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
\label{fig:tc}
\end{figure*}

@@

We need to cache type checking results too.

The key trick is to separate the process into faster binding which constructs dependency
graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
as an over-approximation without full type checking.)

\subsection{Provided types}

\subsection{Type checking functions}

%
\begin{equation*}
\begin{array}{lcll}
\bkndclr{k}&\in&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&\quad(\textit{Node kinds})\\
c&\in&\bot ~|~(\bndclr{v}, \blblclr{l})&\quad(\textit{Call sites})\\
\end{array}
\end{equation*}
%

Binding of

$o.m(\lambda x\rightarrow x)$

\begin{equation*}
\xymatrix{
&\bnd{val}(o)\\
&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
}  
\end{equation*}

Update Figure~\ref{fig:loop} accordingly

We show cases (2) and (7); cases (1) and (6) that reuse nodes are analogous;
other cases just call bind recursively with no node

\subsection{Type checking over dependency graphs}

\begin{equation*}
\begin{array}{lcll}
\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ (m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n) &\quad(\textit{Types})\\
\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
\end{array}
\end{equation*}

\newpage
~
\newpage

@@

>>>>

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{strclr}{rgb}{0.4,0.4,0.0}
\definecolor{rstrclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.6,0.0,0.2}
\newcommand{\vect}[1]{\langl #1 \rangl}
\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\ball}[1]{\FPeval{\result}{clip(201+#1)}\textnormal{\ding{\result}}}
\newcommand{\lsep}{~\,|\,~}
\newcommand{\num}[1]{\textcolor{numclr}{#1}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\rstr}[1]{\textnormal{\textcolor{rstrclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\qident}[1]{\textnormal{\sffamily \guillemotleft #1\guillemotright}}
\newcommand{\dom}{\ident{dom}}
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}

@@

\section{Formalising live coding infrastructure}
\label{sec:formal}

In this section, we present a formalisation of a live coding infrastructure for a small,
expression-based programming language that supports \kvd{let} binding, member invocations
and $\lambda$ abstractions. This is the necessary minimum for data exploration as described
in the previous section.

@@

The evaluation of function values is similar, but requires three cases. If the body can 
be reduced to a value with no unbound variables (\emph{fun-val}), we return a lambda function that
returns the value. If the body requires only the bound variable (\emph{fun-bind}), we return a 
lambda function with the delayed preview as the body. If the body requires further variables,
the result is a delayed preview.

\paragraph{Caching previews.}
For simplicity, the relation $\Downarrow$ in Figure~\ref{fig:eval} does not specify how previews 
are cached and linked to graph nodes. In practice, this is done by maintaining a lookup table
from graph nodes $\bdnclr{v}$ to (possibly delayed) previews $p$.
Whenever $\Downarrow$ is used to obtain a preview for a graph node, we first 
attempt to find an already evaluated preview using the lookup table. If the preview has not
been previously evaluated, we evaluate it and add it to the lookup table.

@@

\begin{figure*}[t]
\vspace{-0.5em}
\begin{equation*}
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta, c}(e_0.m(e_1, \ldots, e_n)) = &(2b)\\ 
\qquad \bndclr{v}, (\{\bndclr{v}\} \cup V_0 \cup \ldots \cup V_n, E \cup E_0 \cup \ldots \cup E_n)\\
\quad \textnormal{when}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma, \Delta, \bot}(e_0)\\
\quad \textnormal{and}~c_i = (\bndclr{v_0}, \blbl{callsite}(m,i))\hspace{4.8em} (i \in 1\ldots n)\\
\quad \textnormal{and}~\bndclr{v_i}, (V_i, E_i) = \ident{bind}_{\Gamma, \Delta, c_i}(e_i)\qquad (i\in 1\ldots n)\\
\quad \textnormal{and}~(\bknd{mem}(m),[(\bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v_n}, \blbl{arg}(n))]) \notin \ident{dom}(\Delta)\hspace{-200em}.\\
\quad \textnormal{let}~\bndclr{v} = \bnd{mem}(m, s), s~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v}, \bndclr{v_0}, \blbl{arg}(0)), \ldots, (\bndclr{v},\bndclr{v_n}, \blbl{arg}(n)) \}
\end{array}
\qquad\qquad
\begin{array}{ll}
\ident{bind}_{\Gamma, \Delta, (\bndclr{v_c}, \blblclr{l_c} )}(\lambda x \rightarrow e) = \bndclr{v}, (\{\bndclr{v}\} \cup V_0, E \cup E_0)&(7b)\\
\quad \textnormal{when}~(\bknd{var}(x),[(\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v_x} = \bnd{var}(x, s_x), s_x~\textnormal{fresh}\\
\quad \textnormal{and}~\Gamma_1 = \Gamma \cup \{ x, \bndclr{v_x} \}\\
\quad \textnormal{and}~\bndclr{v_0}, (V_0, E_0) = \ident{bind}_{\Gamma_1, \Delta, \bot}(e)\\
\quad \textnormal{when}~(\bknd{fun}(x),[(\bndclr{v_0}, \blbl{body}), (\bndclr{v_c}, \blblclr{l_c} )])\notin \ident{dom}(\Delta)\\
\quad \textnormal{let}~\bndclr{v} = \bnd{fun}(x, s_f), s_f~\textnormal{fresh}\\
\quad \textnormal{let}~E = \{ (\bndclr{v},\bndclr{v_0},\blbl{body}), (\bndclr{v},\bndclr{v_c},\blblclr{l_c}), (\bndclr{v_x},\bndclr{v_c},\blblclr{l_c}) \}
\\[0.75em]
\end{array}
\end{equation*}
\vspace{-0.5em}
\caption{Revised binding rules, tracking call sites of function values.} 
\label{fig:binding-rules-callsite}
\vspace{-0.5em}
\end{figure*}

Evaluating live previews can be an expensive operations, so being able to cache partial previews
is a must for a live coding environment. Type checking is typically fast, so the main focus of
this paper is on live previews. However, asynchronous type providers in The Gamma (Section~\ref{sec:types-providers}) 
can make type checking time consuming, and so we use the dependency graph also for type checking
(Section~\ref{sec:types-graph}). Type checking lambda functions (Section~\ref{sec:types-funs}) 
requires a slight extension of the model discussed in Section~\ref{sec:formal}.

@@

\subsection{Asynchronously provided types}
\label{sec:types-providers}

Data available in The Gamma can be defined using several kinds of type providers. The type provider
used in Figure~\ref{fig:thegamma} as asynchronous [X]. It downloads the sample URL and generates 
types based on the contents of the web page. The parameter to \ident{web.scrape} is a static 
parameter and is evaluated during type-checking. We omit details in this paper, but we note this
works similarly to F\# type providers [X].

Type providers can also be implemented as REST services [X] to allow anyone implement a data source
in the language of their choice. In this case, each member of a call-chain returns a type that is
generated based on the result of an HTTP request. For example, when the user types \ident{worldbank}
(to access information about countries), the type provider makes a request to   
\url{http://thegamma-services.azurewebsites.net/worldbank}, which returns two members:
%
\begin{equation*}
\begin{array}{l}
[~\{\rstr{name}\!:\str{byYear},\rstr{returns}\!:\\
~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickYear}\}\}\\
\hspace{0.55em}\{\rstr{name}\!:\str{byCountry},\rstr{returns}\!:\\
~\quad \{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickCountry}\}\}~]
\end{array}
\end{equation*}
%
This indicates that \ident{worldbank} has members \ident{byYear} and \ident{byCountry}.
If the user types \ident{worldbank.byCountry}, a request is made to the specified URL 
\url{http://thegamma-services.azurewebsites.net/worldbank/pickCountry}:
%
\begin{equation*}
\begin{array}{l}
[~\{\rstr{name}\!:\str{Andorra},\rstr{trace}\!:[\str{"country=AR"}], \\
\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\}\\
\hspace{0.55em}\{\rstr{name}\!:\str{Afghanistan},\rstr{trace}\!:[\str{"country=AF"}],\\
\quad \rstr{returns}\!:\{\rstr{kind}\!:\str{nested},\rstr{endpoint}\!:\str{/pickTopic}\}\},\,..]
\end{array}
\end{equation*}
%
This returns a list of countries which can then be accessed as members via 
\ident{worldbank.byCountry.Andorra}, etc.

This is one reason for why type checking in The Gamma can be time consuming. Other
type providers may perform other more computationally intensive work to provide types and so
it is desirable to reuse type-checking results during live coding. The rest of this section
shows how this is done using the dependency graph discussed in Section~\ref{sec:formal}.

\begin{figure}[!b]
\vspace{-0.5em}
\begin{equation*}
\xymatrix{
&\bnd{val}(o)\\
&\bnd{mem}(m, s_0) \ar[u]_{\blbl{arg}(0)} \ar[dl]^{\blbl{arg}(1)} \\
\bnd{fun}(x, s_2) \ar@/^/[uur]^{\blbl{callsite}(m, 1)} \ar[rr]^{\blbl{body}} && \bnd{var}(x, s_1) \ar@/_/[uul]_{\blbl{callsite}(m, 1)}
}  
\end{equation*}
\caption{Dependency graph for $o.m(\lambda x\rightarrow x)$}
\label{fig:graph-func}
\end{figure}

\subsection{Revised binding of functions}
\label{sec:types-funs}

The Gamma script supports lambda functions, but only in a limited way. A function can be passed
as a parameter to a method, which makes type checking of functions easier. For example, consider:
%
\begin{equation*}
\begin{array}{l}  
\ident{movies}.\ident{sortBy}(\lambda x \rightarrow x.\ident{getBudget}())
\end{array}
\end{equation*}
%
If \ident{movies} is a collection of \ident{Movie} objects, the type of the lambda function must be
$\ident{Movie}\rightarrow\ident{bool}$ and so the type of $x$ is \ident{Movie}. This is similar to
type checking of lambda functions in C\# [X], where type is also inferred from the context (or has
to be specified explicitly).

We do not currently allow lambda functions as stand-alone let-bound 
values. This could be done by requiring explicit types, or introducing polymorphism, but it was not
necessary for the limited domain of non-expert data exploration.

\paragraph{Dependency graph for functions.}
In the binding process specified in Section~\ref{sec:formal}, a variable is a leaf of the dependency
graph. In the revised model, it depends on the context in which it appears. A new edge labelled
$\blbl{callsite}(m, i)$ indicates that the source node is the input variable of a function 
passed as the $i^\textnormal{th}$ argument to the $m$ member of the expression represented by the
target node. A node representing function is linked to the call site using the same edge.

Figure~\ref{fig:graph-func} shows the result of binding $o.m(\lambda x\rightarrow x)$. Both
$\bnd{fun}(x, s_2)$ and $\bnd{var}(x, s_1)$ now depend on the node representing $o$. The new 
$\blbl{callsite}$ edge makes it possible to type-check function and variable nodes just using their 
dependencies. As before, the member invocation $\bnd{mem}(m, s_0)$ 
depends on the instance using $\blbl{arg}(0)$ and on its argument using $\blbl{arg}(1)$.

\begin{figure*}
\vspace{0.5em}
\begin{equation*}
\begin{array}{l}
\qquad\inference[(val)~]
  {\Sigma(n) = \alpha}
  {\bnd{val}(n) \vdash \alpha }
\quad\qquad
\inference[(var)~]
  { (\bnd{var}(x,s), \bndclr{v}, \blbl{callsite}(m, i)) \in E &
    \bndclr{v} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) & \tau_i=\tau'\rightarrow\tau'' }
  { \bnd{var}(x, s) \vdash \tau' }
\\[2.5em]
\inference[(fun)~]
  { \{(\bnd{fun}(x,s), \bndclr{v_b}, \blbl{body}), (\bnd{var}(x,s), \bndclr{v_c}, \blbl{callsite}(m, i))\} \subseteq E &
    \bndclr{v_c} \vdash (.., m:(\tau_1, \ldots, \tau_k)\rightarrow \tau,..) &
    \tau_i=\tau'\rightarrow\tau'' & \bndclr{v_b} \vdash \tau'' }
  { \bnd{fun}(x, s) \vdash \tau' \rightarrow \tau'' }
\\[2.5em]
\qquad\qquad\quad\inference[(mem)~]
  { \forall i\in\{0\ldots k\}.(\bnd{mem}(m, s), \bndclr{v_i}, \blbl{arg}(i)) \in E &
    \bndclr{v_0} \vdash (.., m:(\tau_1, \ldots, \tau_k) \rightarrow \tau, ..)& \bndclr{v_i} \vdash \tau_i }
  { \bnd{mem}(m, s) \vdash \tau }
\\[1em]
\end{array}
\end{equation*}
\caption{Rules that define evaluation of previews over a dependency graph for an expression}  
\vspace{0.5em}
\label{fig:tc}
\end{figure*}

\paragraph{Revised binding process.}
For the revised binding process, we introduce a new edge label $\blbl{callsite}$. Variable nodes
now have dependencies and so we cache them and attach a symbol $s$ to \bnd{var}, so we also 
introduce a node kind $\bknd{var}$ as part of a lookup key for $\Delta$.

The \ident{bind} function now has a parameter $c$, in addition to $\Gamma$ and $\Delta$, which 
represents the context in which the binding happens. This is either a member invocation (labelled 
with instance node $\bndclr{v}$ and \blbl{callsite} label $\blblclr{l}$), or not a member 
invocation writted as $\bot$. The updated definitions are:
%
\begin{equation*}
\begin{array}{lcll}
\bndclr{v} &\hspace{-0.5em}\in\hspace{-0.5em}&\bnd{val}(n)~|~\bnd{var}(x, s)~|~\bnd{mem}(m, s)~|~\bnd{fun}(x, s)&(\textit{Vertices})\\
\blblclr{l}&\hspace{-0.5em}\in\hspace{-0.5em}&\blbl{body}~|~\blbl{arg}(i)~|~\blbl{callsite}(m,i)&(\textit{Edge labels})\\
\bkndclr{k}&\hspace{-0.5em}\in\hspace{-0.5em}&\bknd{fun}(x)~|~\bknd{mem}(m)~|~\bknd{var}(x)&(\textit{Node kinds})\\
          c&\hspace{-0.5em}\in\hspace{-0.5em}&\bot ~|~(\bndclr{v}, \blblclr{l})&(\textit{Call sites})\\
\end{array}
\end{equation*}
%
The key parts of the revised definiton of the \ident{bind} function are shown in Figure~\ref{fig:binding-rules-callsite}.
We now write $\ident{bind}_{\Gamma,\Delta,c}$ where $c$ represents the context in which the
binding occurs. This is set to $\bot$ in all cases, except when binding arguments of a member call.
In (2b), we first recursively bind the instance node using $\bot$ as the context and then 
bind all the arguments using $(\bndclr{v_0}, \blbl{callsite}(m, i))$ as the context for $i^{\textnormal{th}}$ 
argument. The rest is as in case (2) before. The case (1) is updated similarly and is not shown 
here for brevity.

When binding a function (7b), we now also store variable nodes in $\Delta$ and so we check if a 
variable with the given call site exists. If no, we create a fresh node $\bnd{var}(x, s_x)$.
The node is added to $\Gamma$ as before. At the end, we now also include a call site edge from the 
variable node and from the function node in $E$. We omit a similar variant of the case (6).
The remaining cases (3)-(5) are the same, except that \ident{bind} has the additional $c$ parameter
and recursive calls always set it to $\bot$.

Finally, the \ident{update} function in Figure~\ref{fig:loop} also needs to be updated to store the
newly created \bnd{var} nodes. This is done by adding the following single case:

\begin{equation*}
\begin{array}{l}
\Delta_{i}(\bknd{var}(x), [(\bndclr{v},\blbl{callsite}(m,i))]) = \bnd{var}(x, s)\\
\quad\quad \textnormal{for all}~\bnd{var}(x, s) \in V\\
\quad\quad \textnormal{such that}~(\bnd{var}(x, s), \bndclr{v}, \blbl{callsite}(m,i)) \in E
\end{array}
\end{equation*}

\subsection{Type checking over dependency graphs}
\label{sec:types-graph}

The type system for The Gamma supports a number of primitive types (such as integers and strings) 
written as $\alpha$. Composed types include functions and objects with members. Objects are be 
provided by type providers, but we omit the details here. Types of object members are written as
$\sigma$ and can have multiple arguments:
%
\begin{equation*}
\begin{array}{lcll}
\tau&\in&\alpha ~|~ \tau \rightarrow \tau ~|~ \{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\} &\quad(\textit{Types})\\
\sigma &\in& (\tau_1, \ldots, \tau_n) \rightarrow \tau &\quad(\textit{Members})\\
\end{array}
\end{equation*}
%
The typing judgements are written in the form $\bnd{v}\vdash \tau$. They are parameterised by the
dependency graph $(V, E)$, but this is not modified during type checking so we keep it implicit
rather than writing e.g.~$\bnd{v}\vdash_{(V,E)} \tau$.

\paragraph{Type checking.}
The typing rules are shown in Figure~\ref{fig:tc}. Types of primitive values $n$ are obtained
using a global lookup table $\Sigma$ (\emph{val}). When type checking a member call (\emph{mem}),
we find its dependencies $\bndclr{v_i}$ and check that the first one (instance) is an object with 
the required member $m$. The types of input arguments of the member then need to match the types 
of the remaining (non-instance) nodes.

Type checking a function (\emph{fun}) and a variable (\emph{var}) is similar. In both cases,
we follow the \blbl{callsite} edge to find the member that accepts the function as an argument.
We obtain the type of the function from the type of the $i^\textnormal{th}$ argument of the 
member. We use the input type as the type of variable (\emph{var}). For functions, we also check 
that the resulting type matches the type of the body (\emph{fun}).

\paragraph{Caching results.}
Performing type checking over the dependency graph, rather than over the abstract syntax tree,
enables us to reuse the results of previously type checked parts of a program. As when 
caching evaluated previews (Section~\ref{sec:previews}), we build a lookup table mapping graph
nodes to types. When type checking a node, we first check the cache and, only if it is new, 
follow the $\vdash$ relation to obtain the type.

As a result, code can be type checked on-the-fly during editing, even when asynchronous type 
providers are used, and the programmer gets instant feedback without delays.

@@

\section{Summary}
The key trick is to separate the process into faster binding which constructs dependency
graph and slower type checking and evaluation. (This assumes that binding can be done, perhaps
as an over-approximation without full type checking.)
